---
title: "Finelli & linear programming"
author: "Eija-Leena Koponen"
date: "`r format(Sys.Date())`"
output:
  html_document:
    keep_md: true
---

### Optimizing

https://fineli.fi/fineli/fi/elintarvikkeet 

FIN:
Mahdollisimman vähäsokerinen 7 pv:n ruokavalioehdotus seuraavilla spekseillä: 
- kokonaisuus mahdollisimman tarkkaan 2000 kcal/pv
- energiasta 50% hiilareista, 30% proteiineista, 20% rasvoista mahdollisimman tarkkaan
- kuitua vähintään 20 g/pv 
- ei samasta ruokaryhmästä (esim. “ateriasalaatti” tai “hampurilainen”) peräkkäisinä päivinä
- vähintään 5 eri ruoka-ainetta per päivä

ENG:
Lowest possible amount of sugar in the 7 days diet with following conditions:
- energy intake 2000 kcal/day
- of which 50 % should be carbohydrates, 30% protein, 20% fat
- at least 20g/day fibers
- on the following days, no items from the same foodgroup
- at least 5 different items per day

```{r eval=TRUE, warning=FALSE, message = FALSE}
library(readr)
library(dplyr)
library(tidyverse)

options(encoding = "UTF-8", scipen = 999)
Sys.getlocale()
#Sys.setlocale("fi_FI.UTF-8/fi_FI.UTF-8/fi_FI.UTF-8/C/fi_FI.UTF-8/fi_FI.UTF-8") #ääkköset

elintarvikkeet <- read_delim("https://fineli.fi/fineli/fi/elintarvikkeet/resultset.csv", 
        ";", escape_double = FALSE, 
        locale = locale("fi", decimal_mark=".",
        asciify = TRUE), 
    na = "N/A", trim_ws = TRUE) %>%  
  mutate_all(funs(str_replace(., "<", "")))%>%
  mutate_at(vars(-"name"),  as.numeric)

summary(elintarvikkeet)

```
Remove spaces from names in order to refer to them more easily.
```{r eval=TRUE, warning=FALSE, message = FALSE}
colnames(elintarvikkeet) <- gsub(" ", "_", colnames(elintarvikkeet))
colnames(elintarvikkeet) <- gsub("-", "", colnames(elintarvikkeet))
colnames(elintarvikkeet) <- gsub(",", "", colnames(elintarvikkeet))
colnames(elintarvikkeet)<- gsub("[()]", "", colnames(elintarvikkeet))
```

Select columns needed
```{r eval=TRUE, warning=FALSE, message = FALSE}

elintarvikkeetmod <- elintarvikkeet%>%
  mutate(ruokaryhma=gsub(",.*$", "", name),
         kcal=energia_laskennallinen_kJ*0.239005736,
         sum_hpr=hiilihydraatti_imeytyvä_g+proteiini_g+rasva_g)
  
  
elintarvikkeetmod%>%select(name, ruokaryhma, sokerit_g,
                        energia_laskennallinen_kJ, kcal,
                        hiilihydraatti_imeytyvä_g,
                        proteiini_g,
                        rasva_g,
                        kuitu_kokonais_g, kuitu_veteen_liukenematon_g
                        )%>%summary()

```
```{r eval=TRUE, warning=FALSE, message = FALSE}
library(ggplot2)
library(GGally)

elintarvikkeetmod%>%select(ruokaryhma, sokerit_g,
                        energia_laskennallinen_kJ, kcal,
                        hiilihydraatti_imeytyvä_g,
                        proteiini_g,
                        rasva_g,
                        kuitu_kokonais_g
                        )%>%ggpairs(.
                                    ,columns = 2:8
                                    #,mapping=aes(colour=ruokaryhma)
                                    ) +theme_minimal()

```

And another picture

```{r eval=TRUE, warning=FALSE, message = FALSE}
library(plotly)
selected <- c("Ateriasalaatti", "Hampurilainen")

elintarvikkeetmod%>%select(name, ruokaryhma, sokerit_g,
                        energia_laskennallinen_kJ, kcal,
                        hiilihydraatti_imeytyvä_g,
                        proteiini_g,
                        rasva_g,
                        kuitu_kokonais_g
                        )%>%
  filter(ruokaryhma %in% selected)%>%
   plot_ly(data = ., x = ~kcal, y = ~sokerit_g, color = ~ruokaryhma,
           text = ~name)


elintarvikkeetmod%>%select(name, ruokaryhma, sokerit_g,
                        energia_laskennallinen_kJ, kcal,
                        hiilihydraatti_imeytyvä_g,
                        proteiini_g,
                        rasva_g,
                        kuitu_kokonais_g
                        )%>%
  filter(ruokaryhma %in% selected)%>%
   plot_ly(data = ., x = ~kcal, y = ~proteiini_g, color = ~ruokaryhma,
           text = ~name)
```

The objective function in the above problem will be:

min(sokerit)

when 0.5hiilarit+0.3proteiini+0.2rasvat
2000kcal
min(kuitu)=20g
count(rows)>=5


```{r eval=TRUE, warning=FALSE}
library(lpSolve)
 # Constraint matrix
constraints <- rbind(elintarvikkeetmod$kcal,
                elintarvikkeetmod$kcal, 
                elintarvikkeetmod$kuitu_kokonais_g, 
                nrow(elintarvikkeetmod),
                (elintarvikkeetmod$hiilihydraatti_imeytyvä_g/elintarvikkeetmod$sum_hpr),
                (elintarvikkeetmod$proteiini_g/elintarvikkeetmod$sum_hpr),
                (elintarvikkeetmod$rasva_g/elintarvikkeetmod$sum_hpr)
               
                )

mod <- lp("min",  # min/max
          elintarvikkeetmod$sokerit_g,  # Objective
         constraints   ,  # Constraint matrix
          c(">=", "<=", ">=",">=", ">=",">=",">="),  # Constraint directions
          c(1990,2010, 20,5, 0.5,0.3,0.2), #Limits
          all.bin = TRUE
         ,num.bin.solns = 150
)


mod
```
Results

```{r eval=TRUE, warning=FALSE, message=FALSE}
### solve the problem multiple times
#solved1<- lp("max", f.obj, f.con, f.dir, f.rhs, all.bin=TRUE,   num.bin.solns = 10)
#solved1

numcols <- ncol(constraints) 
numsols <- mod$num.bin.solns

solutions_all <- matrix(head(mod$solution, numcols*numsols), nrow=numsols, byrow=TRUE)


##attach them to the dataframe
for(i in 1:numsols){
  elintarvikkeetmod[, ncol(elintarvikkeetmod) + 1] <- solutions_all[i,]#rnorm(nrow(simple))
  names(elintarvikkeetmod)[ncol(elintarvikkeetmod)] <- paste0("solution", i)
}

```

Check results

```{r}

library(tidyr)
library(reshape2)

solutionstacked<- elintarvikkeetmod%>%
  select(ruokaryhma, name, kcal,sokerit_g,kuitu_kokonais_g, solution1:paste0("solution",numsols))%>%
  reshape2::melt(., id=c("ruokaryhma","name","kcal", "sokerit_g", "kuitu_kokonais_g"))%>%
  filter(value==1)

solutionstacked%>%
  group_by(variable)%>%
  summarise(sum_kcal=sum(kcal),
            sum_sokerit=sum(sokerit_g),
            sum_kuitu=sum(kuitu_kokonais_g))%>%
  head()

```


Select 7 day diet from the options, no same ruokaryhmä on consecutive days
```{r eval=TRUE, warning=FALSE, message = FALSE}
library(tidyr)
library(reshape)


solutionstackedmatrix <- solutionstacked%>%
  select(ruokaryhma, name, variable, value)%>%
  reshape::cast(., ruokaryhma+name ~ variable, mean)

firstselection <- elintarvikkeetmod%>%
  select(solution1:paste0("solution",numsols))%>%
  names()%>%
  sample(., 1, replace = FALSE, prob = NULL)

discard <- solutionstacked%>%filter(variable==firstselection)%>%
  group_by(variable, ruokaryhma)%>%summarise(n=n())

nextselectiondiscard <-solutionstacked %>%
    group_by(variable) %>%
    filter(!any(ruokaryhma %in% discard$ruokaryhma))%>%
    select(variable)%>%as.list()

nextselectiondiscard
##doesn't give any other 'ruokaryhmä' :(

#nextselection<- 


```
With these results /approach it is not possible to select from different food groups!
